---
title: "Computational Statistics Lab 1 Report Group 6"
author: Jaskirat S Marar (jasma356), Filip Berndtsson (filbe354), Dinuke Jayaweera (dinja628), Raja Uzair Saeed (rajsa233), Mucahit Sahin (mucsa806)
date: "11/9/2021"
output: pdf_document
---

```{r setup, include=FALSE}
options(digits = 22)
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

## 1. Check the results of the snippet. Comment on whats going on

First Snippet:
```{r}
x1 <-  1 / 3 ;
x2 <- 1 / 4
if (x1 - x2==1/12){
  print("Subtraction is correct") }else{print("Subtraction is wrong")
  }
```





## 2. If there are any problems suggest solutions

# Problem 2

## 1. Write your own R function to calculate the derivative of f(x) = x in this way with $\epsilon$ = 10^-15^

## 2. Evaluate your derivative function at x = 1 and x = 100000

## 3. What values did you obtain? What are the true values? Explain the reasons behind the discovered differences

# Problem 3

## 1. Write your own R function, myvar, to estimate the variance in this way

## 2. Generate a vector x = (x~1~,..., x~10000~) with 10000 random numbers with mean 10^8^ and variance 1

## 3. For each subset X~*i*~ = {x~1~, ... , x~*i*~}, i = 1,..., 10000 compute the difference Y~*i*~ = myvar(X~*i*~)-var(X~*i*~), where var(X~*i*~) is the standard variance estimation function in R. Plot the dependence Y~*i*~ on *i*. Draw conclusions from this plot. How well does your function work? Can you explain the behaviour?

## 4. How can you better implement a variance estimator? Find and implement a formula that will give the same results as var()?

# Problem 4

## Consider the three below R expressions for computing the binomial coefficient. They all use the prod() function, which computes the product of all the elements of the vector passed to it.
## A) prod ( 1 : n) / (prod ( 1 : k ) * prod ( 1 : ( n-k ) ) )
## B) prod( ( k+1) : n) / prod ( 1 : ( n-k ) )
## C) prod ( ( ( k+1) : n) / ( 1 : ( n-k ) ) )

## 1. Even if overflow and underflow would not occur these expressions will not work correctly for all values of n and k. Explain what is the problem in A, B and C respectively

## 2. In mathematical formulae one should suspect overflow to occur when parameters, here n and k, are large. Experiment numerically with the code of A, B and C, for different values of n and k to see whether overflow occurs. Graphically present the results of your experiments.

## 3. Which of the three expressions have the overflow problem? Explain why.